Cartogram Mapping and its Application to Cancer Data Visualisation

### 2019-08-22 review_article.Rmd
Supporting statistics

The visualisation methods used to present cancer statistics will depend on the intended message and users of the map. 
Transforming statistics to visualisations takes advantage of the human visual system, which has been demonstrated to an effective communication channel for a wide range of audiences (insert munzer reference). Visualising cancer statistics enables the non-expert audience to access the outputs of complicated spatial epidemeaological analyses. These visualisation show individual observations, or observation points, aggregated into communities and geographical units.
Visualising diseases on maps is often the first step in exploratory spatial data analysis and effectively helps in the formulation of hypotheses [@MTMSIH]. 
Disease maps help to present geographic patterns that may be overlooked in a table, obscuring the geospatially related statistics [@SAMGIS].
By providing a visual representation of cancer statistics, geographic patterns of disease are able to be identified, investigated further and/or effectively addressed.
@SE recognises one of the key challenges with mapping spatial patterns of disease is the design of visualisations.
This paper addresses the visualisation techniques and their applications to cancer statistics. Highlighting the differences and historic use of these displays.
Transforming statistics to visualisations takes advantage of the human visual system, which has been demonstrated to an effective communication channel for a wide range of audiences (insert munzer reference). Visualising cancer statistics enables the non-expert audience to access the outputs of complicated spatial epidemeological analyses. These visualisation show individual observations, or observation points, aggregated into communities and geographical units.
Visualising diseases on maps is often the first step in exploratory spatial data analysis and effectively helps in the formulation of hypotheses [@MTMSIH]. 
Disease maps help to present geographic patterns that may have been overlooked in a table, obscuring the geospatially related statistics [@SAMGIS].
By providing a visual representation of cancer statistics, geographic patterns of disease are able to be identified, nvestigated further and/or effectively addressed.

# Disease mapping methods
<!-- insert paragraph summarising/ stating the different types of maps used for cancer mapping. --->

@@

<!-- what is a choropleth map -->
A choropleth map is used to display the characteristics of a spatial relationship of measurements. 
A choropleth map is a true map of the topology, constructed for visual inspection of spatial patterns on a familiar geographic form. 
Figure 1 shows a choropleth map where each state of the United States of America is coloured by the average annual rate of new cases of lung and bronchus cancer from years 2012 to 2016.
Identifying and explaining spatial structures, patterns, and processes involves considering the individuals in communities and organising communities into representable units [@SAMGIS]. 
A choropleth is constructed by drawing the geographic or political boundaries, and filling the shapes with colours to represent values of a measured variable [@EI].
Early versions of choropleth maps used symbols or patterns instead of colour.

As an alternative storage device to a table, it preserves locations for geographically ordered data, with usage dating back to the 1800s [@GOINO].
@CPISACA discuss the use of choropleths to visualise cancer data, and @DMAHP gives an overview of the development of these maps for displaying disease data.
@CPISACA discusses their use in visualising cancer data, and @DMAHP gives an overview of the development of these maps for displaying disease data.
Just as geographers are no longer the only creators of maps, @CPISACA suggests the audiences of spatial health data analysis have extended beyond researchers to the public, policymakers and the media. As such, many examples of choropleth maps are found on the websites of government, media, health and other public facing organisations (insert ref for Jessie's grey lit review).
A choropleth map is a true map of the topology, constructed for visual inspection of spatial patterns across a familiar geographic form. 
Figure 1 shows a choropleth map where each state of the United States of America is coloured by the average annual rate of new cases of lung and bronchus cancer from years 2012 to 2016.

@@

<!-- Measures reported in cancer atlases -->
Almost 100 years of cancer mapping in the United States and the United Kingdom has seen increased effectiveness in the presentation cancer statistics.
Mortality rates are now often presented as relative rates of risk across the population, and age adjusted to correct for the higher prevalence of cancers in older populations.

Howe [-@HEDP] describes Stock's development of the standardised mortality ratios through the 1930s. Table \ref{measures} summarises the measures presented in published cancer atlases, between 2005 and 2015 and provides a definition of each measure.

@@

```{r choropleth-grid-create, echo=FALSE, message=FALSE, eval = FALSE,}
library(grid)
library(png)
library(gridExtra)

plot1 <- png::readPNG('cancer_map_images/choropleths/15.choropleth.png')
plot2 <- png::readPNG('cancer_map_images/choropleths/25.choropleth.png')
plot3 <- png::readPNG('cancer_map_images/choropleths/1.png')
plot4 <- png::readPNG('cancer_map_images/choropleths/29.choropleth.png')
plot5 <- png::readPNG('cancer_map_images/choropleths/30.choropleth.png')
plot6 <- png::readPNG('cancer_map_images/choropleths/6.choropleth.png')
plot7 <- png::readPNG('cancer_map_images/choropleths/8.choropleth2.png')
plot8 <- png::readPNG('cancer_map_images/choropleths/20.choropleth1.png')
plot9 <- png::readPNG('cancer_map_images/choropleths/7.choropleth.png')
plot10 <- png::readPNG('cancer_map_images/choropleths/23.choropleth.png')
plot11 <- png::readPNG('cancer_map_images/choropleths/24.choropleth.png')
plot12 <- png::readPNG('cancer_map_images/choropleths/14.choropleth1.png')
plot13 <- png::readPNG('cancer_map_images/choropleths/26.choropleth.png')
plot14 <- png::readPNG('cancer_map_images/choropleths/27.choropleth.png')
plot15 <- png::readPNG('cancer_map_images/choropleths/21.choropleth2.png')
plot16 <- png::readPNG('cancer_map_images/choropleths/3.choropleth.png')
plot17 <- png::readPNG('cancer_map_images/choropleths/5.choropleth.png')

choropleth_grid <- gridExtra::grid.arrange(
  grid::rasterGrob(plot1), 
  grid::rasterGrob(plot2), 
  grid::rasterGrob(plot3), 
  grid::rasterGrob(plot4), 
  grid::rasterGrob(plot5), 
  grid::rasterGrob(plot6),
  grid::rasterGrob(plot7), 
  grid::rasterGrob(plot8), 
  grid::rasterGrob(plot9),
  grid::rasterGrob(plot10),
  grid::rasterGrob(plot11), 
  grid::rasterGrob(plot12),
  grid::rasterGrob(plot13), 
  grid::rasterGrob(plot14), 
  grid::rasterGrob(plot15),
  grid::rasterGrob(plot16),
  grid::rasterGrob(plot17))

ggsave(choropleth_grid, filename = "figures/choropleth_grid.png", device = "png", width = 12, height = 6)

```

```{r choropleth-grid-print, echo = FALSE, out.width= '100%', fig.cap = "Insert grid title and references to individual tiles (ref)"}
knitr::include_graphics("figures/choropleth_grid.png")
```

@@

The atlases provided additional information to help users understand and interpret the spatial distribution presented. Table \ref{tab:supporting-stats} explores the use of these supports in a variety of online map sources. Demographics include information regarding the age and sex distribution of the areas presented. Socioeconomic indicators include Unemployment rates of regions, Remoteness and Education levels achieved. Statistical uncertainty is communicated through a Confidence (CI) interval, statistical significance levels, Boxplots, Distribution plots, and reporting the Sample Size and Standard Deviations.
These additional pieces of information can help to validate, explain or explore relationships between cancer incidence and survival rates.

Table: \label{tab:supporting-stats} Supporting statistics in public facing atlases

| Atlas                    | Geographic hierarchies | Population Distributions | Demographics | Statistical Uncertainty | Socioeconomic Indicators |
|--------------------------|------------------------|--------------------------|--------------|-------------------------|--------------------------|
| All Ireland Cancer Atlas |                        |                          |              |                         |                          |
|                          |                        |                          |              |                         |                          |
|                          |                        |                          |              |                         |                          |

<!-- geographic hierarchies  -->
While atlases are often used to describe differences between areas, the may allow statistics to be displayed at different levels of aggregation. World atlases can allow for displays of data aggregated into continents, country, states or provinces [@IARC_3].

<!-- population distribution -->
It is likely that each population area will have a different amount of people, information regarding the distribution of population levels may be provided in a table or histogram display [@NICR_1]. Atlases could also connect the population to the land available to them by communicating population density.

<!-- statistical uncertainty -->
These additional statistics often include a measure of the statistical uncertainty of the values of the statistics presented in a choropleth.
In the review of atlases in the public domain, atlases were considered to report uncertainty to the non-expert user if they included a measure of statistical uncertainty either within or alongside the map. The maps considered used standard and well known measures including credible intervals and standard deviation, statistical significance, box plots and distributions. The maps employing uncertainty ranged from static pdfs or infographics, to interactive online resources.
Close to half of the atlases identified (42%, n=14) included some measure of uncertainty. The most common measure used to represent uncertainty were credible or confidence intervals (CIs).

<!-- Demographics -->
Atlases often specified their use of gender in the diplays. Atlases that allowed for inteaction also allowed users to select between displays for males, females or both depending on the type of cancers explored.
<!-- Socioeconomic Indicators -->
Few atlases also explored the impact of rurality on cancer rates.

@@

<!-- @CPISACA contrasted several online atlases in a 2006 paper, consider for structure?

(only worth including if it is possible for us to implement)
Tabular form comparing and contrasting
- Relationship to geography
- Show using cancer examples
-->

>>>>

The visualisation methods used to present cancer statistics will depend on the intended message and users of the map. 
Presenting statistics as visualisations requires transforming individual observations into aggregations of communities as geographical units.
Visualising diseases on maps is often the first step in exploratory spatial data analysis and effectively helps in the formulation of hypotheses [@MTMSIH]. 
Disease maps help to present geographic patterns that may have been overlooked in a table, obscuring the geospatially related statistics [@SAMGIS].
By providing a visual representation of cancer outcomes, geographic patterns of disease are able to be identified and effectively addressed.
@SE recognises one of the key challenges with mapping spatial patterns of disease is the design of visualisations.
This paper addresses the visualisation techniques and their applications to cancer statistics. Highlighting the differences and historic use of these displays.

# Disease mapping methods

@@

<!-- what is a choropleth map -->
A choropleth map is used to display the characteristics of a spatial relationship of measurements. 
A choropleth map is a true map of the topology, constructed for visual inspection of spatial patterns on a familiar geographic form. 
\ref{fig:choro} 1 shows a choropleth map where each state of the United States of America is coloured by the average annual rate of new cases of lung and bronchus cancer from years 2012 to 2016.
Identifying and explaining spatial structures, patterns, and processes involves considering the individuals in communities and organising communities into representable units [@SAMGIS]. 
A choropleth is constructed by drawing the geographic or political boundaries, and filling the shapes with colours to represent values of a measured variable [@EI].
Early versions of choropleth maps used symbols or patterns instead of colour.
As an alternative storage device to a table, it preserves locations for geographically ordered data, with usage dating back to the 1800s [@GOINO].
@CPISACA discuss the use of choropleths to visualise cancer data, and @DMAHP gives an overview of the development of these maps for displaying disease data.

@@

<!-- Measures reported in cancer atlases -->
Almost 100 years of cancer mapping in the United States and the United Kingdom has seen increased effectiveness in the presentation cancer statistics.
Mortality rates are now often presented as relative rates of risk across the population, and age adjusted to correct for the higher prevalence of cancers in older populations.
Howe [-@HEDP] describes Stock's development of the standardised mortality ratios through the 1930s. Table \ref{tab:measures} presents summarises the measures presented in published cancer atlases, and provides a definition of each measure.

@@

```{r choropleth-grid-create, message=FALSE, eval = FALSE}
library(grid)
library(png)
library(gridExtra)

plot1 <- png::readPNG('cancer_map_images/choropleths/1.png')
plot2 <- png::readPNG('cancer_map_images/choropleths/14.choropleth1.png')
plot3 <- png::readPNG('cancer_map_images/choropleths/15.choropleth.png')
plot4 <- png::readPNG('cancer_map_images/choropleths/3.choropleth.png')
plot5 <- png::readPNG('cancer_map_images/choropleths/5.choropleth.png')
plot6 <- png::readPNG('cancer_map_images/choropleths/6.choropleth.png')
plot6 <- png::readPNG('cancer_map_images/choropleths/7.choropleth.png')
plot7 <- png::readPNG('cancer_map_images/choropleths/8.choropleth.png')
plot8 <- png::readPNG('cancer_map_images/choropleths/20.choropleth1.png')
plot9 <- png::readPNG('cancer_map_images/choropleths/21.choropleth2.png')
plot10 <- png::readPNG('cancer_map_images/choropleths/23.choropleth.png')
plot11 <- png::readPNG('cancer_map_images/choropleths/24.choropleth.png')
plot12 <- png::readPNG('cancer_map_images/choropleths/25.choropleth.png')
plot13 <- png::readPNG('cancer_map_images/choropleths/26.choropleth.png')
plot14 <- png::readPNG('cancer_map_images/choropleths/27.choropleth.png')
plot15 <- png::readPNG('cancer_map_images/choropleths/29.choropleth.png')
plot16 <- png::readPNG('cancer_map_images/choropleths/30.choropleth.png')

choropleth_grid <- gridExtra::grid.arrange(grid::rasterGrob(plot1),grid::rasterGrob(plot2), grid::rasterGrob(plot3), 
                        grid::rasterGrob(plot4), grid::rasterGrob(plot5), grid::rasterGrob(plot6),
                        grid::rasterGrob(plot7), grid::rasterGrob(plot8), grid::rasterGrob(plot9),
                        grid::rasterGrob(plot10),grid::rasterGrob(plot11), grid::rasterGrob(plot12), 
                        grid::rasterGrob(plot13), grid::rasterGrob(plot14), grid::rasterGrob(plot15), 
                        grid::rasterGrob(plot16), nrow=4)

ggsave(filename = "figures/choropleth_grid.png", plot = choropleth_grid, device = "png", width = 6, height = 5, dpi = 300)
```

```{r choropleth-grid-print, echo = FALSE, results = "asis"}
knitr::include_graphics("figures/choropleth_grid.png")
```

@@

The atlases provided additional information to help users understand and interpret the spatial distribution presented. The use of these supports were found in a variety of online map sources.
These additional pieces of information can help to validate, explain or explore relationships between cancer incidence and survival rates.

#### Geographic hierarchies
While atlases are often used to describe differences between areas, statistics may be displayed at different levels of aggregation. The World Health Organisation and UN Regions can also be used to aggregate areas [@IARC_3].
World atlases can allow for displays of data aggregated into continents, countries, states, provinces and congressional districts [@USCS].

#### Population distribution
It is likely that each population area will have a different amount of people, information regarding the distribution of population levels may be provided in a table or histogram display [@NICR_1]. 
Atlases have the opportunity to connect the population to the land available to them by communicating population density.

#### Statistical uncertainty
Statistical uncertainty is communicated through a Confidence (CI) interval, statistical significance levels, Boxplots, Distribution plots, and reporting the Sample Size and Standard Deviations.
The additional statistics often include a measure of the statistical uncertainty of the values of the statistics presented in a choropleth.
In the review of atlases in the public domain, atlases were considered to report uncertainty to the non-expert user if they included a measure of statistical uncertainty either within or alongside the map. 
The maps considered used standard and well known measures including credible intervals and standard deviation, statistical significance, box plots and distributions.
The maps employing uncertainty ranged from static pdfs or infographics, to interactive online resources.
Close to half of the atlases identified (42%, n=14) included some measure of uncertainty. The most common measure used to represent uncertainty were credible or confidence intervals (CIs).

#### Demographics
Demographics include information regarding the age and sex distribution of the areas presented. 
<!-- sex -->
Sex is an important cofactor for cancer atlases. 
As some cancers are sex specific, and others may be found in both males and females, atlases often specify their use of gender in the diplays. 
Atlases can allow for interaction as users can select between displays for males, females or both depending on the type of cancers explored.

#### Socioeconomic indicators
Socioeconomic indicators can explain how the experience of cancer prevalence varies for various members of a society.
These indicators include unemployment rates, poverty rates and percentages, remoteness and education levels achieved though only a few atlases also explored the impact of rurality on cancer rates. 
The Human Development Index can be used to understand the socioeconomic experience of a community, as can the Income levels as measured by World bank list of economies[@IARC_3].

@@
